---
title: "Advent of Code: 2022 Day 1 in R"
date: 2023-09-07
categories:
  - r
  - advent of code
  - puzzle
---

Below is my solution for Advent of Code 2022 day 1 in R. See the puzzle instructions [here](https://adventofcode.com/2022/day/1).

## Part 1

### Read Input

The first part of any Advent of Code puzzle is reading in the input file. I always copy the example input given to use as a test case. That's [test_input.txt](test_input.txt). I have also copied my actual puzzle input into [input.txt](input.txt).

In this case, we need to read the lines of the input file and find a way to work with the data structure in R. Just reading in the data is very easy:

```{r}
read_input <- function(file) {
  as.numeric(readLines(file))
}
```

That will give us a numeric vector where each line in the file is an element in the vector, and the empty lines that separate the elves will be `NA`. This will actually make it pretty easy to iterate through because we can use `is.na()` to determine when we're moving to a new Elf.

### Solve

The solution is relatively straightforward. We keep a running total of Calories for each Elf, then when we get to the end of that Elf's values we check to see if that Elf's total is higher than the highest we've yet seen.

```{r}
part_1 <- function(input_data) {
  highest_total <- 0 # <1>
  running_total <- 0 # <1>
  for (i in input_data) { # <2>
    if (is.na(i)) { # <3>
      if (running_total > highest_total) { # <3>
        highest_total <- running_total # <3>
      } # <3>
      running_total <- 0 # <4>
      next # <5>
    }
    running_total <- running_total + i # <6>
  }
  if (running_total > highest_total) { # <7>
    highest_total <- running_total # <7>
  }
  highest_total
}
```
1. Initialize variables to use in the loop
2. Iterate over each element in `input_data`
3. If `is.na(i)` we know that we've reached the end of the Elf's values, so we can check the running total against the highest total that we've seen
4. Re-set the `running_total` variable to 0 so that we can start clean with the next Elf
5. We use `next` to move onto the next iteration of the loop so that we don't try to do calculations on `NA` 
6. If `!is.na(i)` then we just add the current element to the running total
7. Finally, we know that checking `is.na(i)` will not capture the elements for the last Elf in the input. So after the loop, we check one last time if `running_total > highest_total`, just to check for that last Elf.

We need to run this with the example data to make sure we come up with 24,000:

```{r}
test_data <- read_input("test_input.txt")
part_1(test_data)
```

Success! Now we try it on the real data.

```{r}
real_data <- read_input("input.txt")
part_1(real_data)
```

Advent of Code tells me that's correct!

## Part 2

### Solve

This one is a bit trickier because we need to keep track of all of the Elves' totals and take the top three once we've counted them all up. We can use the same input data as before and we don't need to mess with the function to read it in. What I will do now, instead of keeping track of the single highest total, is just populate a vector with all of the totals, sort it, and then take the three highest values. I can figure out the size of vector that I need by getting the number of `NA`s and adding one -- there is an `NA` between every Elf, but none after the last one, so that has to be added on.

```{r}
part_2 <- function(input_data) {
  elf_totals <- numeric(sum(is.na(input_data)) + 1) # <1> 
  
  current_elf <- 1 # <2>
  running_total <- 0 # <2>
  for (i in seq_along(input_data)) { # <3>
    if (i == length(input_data)) { # <4>
      running_total <- running_total + input_data[i] # <4>
      elf_totals[current_elf] <- running_total # <4>
    } # <4>
    if (is.na(input_data[i])) { # <5>
      elf_totals[current_elf] <- running_total # <5>
      running_total <- 0 # <5>
      current_elf <- current_elf + 1 # <5>
      next # <5>
    } # <5>
    running_total <- running_total + input_data[i] # <6>
  }
  sum(sort(elf_totals, decreasing = TRUE)[1:3]) # <7>
}
```
1. Pre-allocate the vector that will contain the Elves' totals
2. Initialize variables
3. Iterate through the input data
4. If we're at the end of the data, we need to update the `running_total` and tack that onto `elf_totals`. We have to do this check separately from the `is.na()` check because the data ends with the last Elve's last snack, not `NA`. In retrospect, I could've potentially simplified the code by adding `NA` to the end of the input data vector, but checking manually for the end of the vector isn't particularly hard or confusing, so I just stick with that.
5. If the current value of the vector is `NA`, we need to populate the `elf_totals` vector with the `running_total` that we've been adding up. We then reset `running_total` to 0 and increment the `current_elf` count, before continuing in the loop.
6. Add the current value in the data vector onto the `running_total`
7. Finally, we sort the `elf_totals` vector, specifying `decreasing = TRUE`, take the first three elements, and then return the sum.

Checking this on the example data, making sure it returns 45,000:

```{r}
part_2(test_data)
```

Good! And now the real data:

```{r}
part_2(real_data)
```

Which Advent of Code tells me is correct.

So that's all there is to it! Happy coding!