---
title: "Analysis of a Conjoint Survey Experiment"
date: 2026-04-26
categories:
  - r
  - dplyr
  - data science
execute: 
  message: false
  warning: false
---

Below is the analysis of a conjoint survey experiment that I performed for [Professor Mara Revkin](https://www.mararevkin.com/) at Duke University. 
I am not able to include the details of the survey experiment such as the text of the questions at this time. 

```{r}
suppressPackageStartupMessages(library(lmtest))
suppressPackageStartupMessages(library(sandwich))
suppressPackageStartupMessages(library(tidyverse))
```


In the questionnaire, this analysis covers chapter 10 questions 74-78. There 3 
iterations of the experimental treatment per respondent.
The relevant variables, where `x` is the iteration number 1-3:  

* `I_x_Q_75` -- iteration x, randomization 1
* `I_x_Q_76` -- iteration x, randomization 2
* `I_x_Q_77` -- iteration x, randomization 3
* `I_x_Q_78` -- iteration x, randomization 4
* `I_x_Q6_4` -- iteration x, response 1
* `I_x_Q6_5_A` -- iteration x, response 2

## Data Wrangling

```{r}
df <- read_csv("data/feb9data.csv")
```

For response variable 2, `Q6_5_A`, I am going to:

* convert `-1` to `NA`
* convert `2. No` to 0
* convert `1. Yes` to 1

I use `sapply()` to convert the responses for this variable.

```{r}
convert_q6_5a <- function(x) {
  if (x == "2. No") return(0)
  else if (x == "1. Yes") return(1)
  else return(NA)
}

df$I_1_Q6_5_A <- unname(sapply(df$I_1_Q6_5_A, convert_q6_5a))
df$I_2_Q6_5_A <- unname(sapply(df$I_2_Q6_5_A, convert_q6_5a))
df$I_3_Q6_5_A <- unname(sapply(df$I_3_Q6_5_A, convert_q6_5a))
```

I'm going to select the subset of columns that I want to actually do analysis
on, using *dplyr*'s `ends_with()` select helper:

```{r}
df_sub <- df |> 
  select(SbjNum,
         ends_with("_Q_75"),
         ends_with("_Q_76"),
         ends_with("_Q_77"),
         ends_with("_Q_78"),
         ends_with("_Q6_4"),
         ends_with("_Q6_5_A"))
```

Now we work on reshaping the data using `gather()`. I conduct a few QA checks to
ensure that the reshaping did not result in any data loss.

```{r}
gathered_vars <- df_sub |> 
  select(SbjNum,
         ends_with("_Q_75"),
         ends_with("_Q_76"),
         ends_with("_Q_77"),
         ends_with("_Q_78")) |> 
  gather(variable, value, -SbjNum) 

iterations <- as.numeric(
  str_match(gathered_vars$variable, "I_(\\d)_Q_7.")[, 2]
)
q_num <- str_match(gathered_vars$variable, "I_\\d_(Q_\\d{2})")[, 2]
```

Let's look at `iterations` and `q_num` to make sure they have the right number of
values and the correct distributions:

```{r}
length(iterations) == nrow(gathered_vars)
length(q_num) == nrow(gathered_vars)
```

```{r}
table(iterations)
```

```{r}
table(q_num)
```

Ok, they all have the correct number of items all equally distributed. I now
create a new data frame with the reshaped data.

```{r}
reshaped_vars <- cbind(gathered_vars, iterations, q_num) |> 
  select(-variable) |> 
  spread(q_num, value) |> 
  rename(iteration = iterations)
```

Quality assurance: Randomly select 5 subject IDs and confirm that the values are
the same between the original data frame and the gathered data:

```{r}
qa_subjects <- sample(gathered_vars$SbjNum, 5)

for (subject in qa_subjects) {
  print(paste("Subject:", subject))
  df_subject_original <- filter(df_sub, SbjNum == subject)
  df_subject_gathered <- filter(reshaped_vars, SbjNum == subject)
  
  for (iter in 1:3) {
    # Q_75
    gathered_75 <- df_subject_gathered |> 
      filter(iteration == iter) |> 
      select(Q_75) |> 
      pull()
    print(paste(
      "Iteration", 
      iter,
      "Q_75:", 
      df_subject_original[[paste0("I_", iter, "_Q_75")]] == gathered_75
    ))
    
    # Q_76
    gathered_76 <- df_subject_gathered |> 
      filter(iteration == iter) |> 
      select(Q_76) |> 
      pull()
    print(paste(
      "Iteration",
      iter,
      "Q_76:", 
      df_subject_original[[paste0("I_", iter, "_Q_76")]] == gathered_76
    ))
    
    # Q_77
    gathered_77 <- df_subject_gathered |> 
      filter(iteration == iter) |> 
      select(Q_77) |> 
      pull()
    print(paste(
      "Iteration", 
      iter,
      "Q_77:", 
      df_subject_original[[paste0("I_", iter, "_Q_77")]] == gathered_77
    ))
    
    # Q_78
    gathered_78 <- df_subject_gathered |> 
      filter(iteration == iter) |> 
      select(Q_78) |> 
      pull()
    print(paste(
      "Iteration",
      iter,
      "Q_78:", 
      df_subject_original[[paste0("I_", iter, "_Q_78")]] == gathered_78
    ))
  }
}
```
That all looks good, I feel confident that the reshaping didn't mess up any of
the data points.

Now I need to do the same kind of reshaping on the response variables.

```{r}
gathered_responses <- df_sub |> 
  select(SbjNum,
         matches("I_\\d_Q6_\\d*")
         ) |> 
  gather(variable, value, -SbjNum) 

iterations <- as.numeric(
  str_match(gathered_responses$variable, "I_(\\d)_Q6")[, 2]
)
q_num <- str_match(gathered_responses$variable, "I_\\d_(Q6.+)")[, 2]
```

Ensure correct number of items:

```{r}
length(iterations) == nrow(gathered_responses)
length(q_num) == nrow(gathered_responses)
```

```{r}
table(iterations)
```

```{r}
table(q_num)
```

That all looks good.

```{r}
reshaped_responses <- cbind(gathered_responses, iterations, q_num) |> 
  select(-variable) |> 
  spread(q_num, value) |> 
  rename(iteration = iterations)
```

Repeat the same kind of QA for the response variables:

```{r}
qa_subjects <- sample(reshaped_responses$SbjNum, 5)

for (subject in qa_subjects) {
  print(paste("Subject:", subject))
  df_subject_original <- filter(df_sub, SbjNum == subject)
  df_subject_reshaped <- filter(reshaped_responses, SbjNum == subject)
  
  for (iter in 1:3) {
    # Q6_4
    reshaped_6_4 <- df_subject_reshaped |> 
      filter(iteration == iter) |> 
      select(Q6_4) |> 
      pull()
    original_response <- df_subject_original[[paste0("I_", iter, "_Q6_4")]]
    if (is.na(original_response) && is.na(reshaped_64)) {
      print(paste("Iteration", iter, "Q6_4:", TRUE))
    } else {
      print(paste(
        "Iteration",
        iter,
        "Q6_4:",
        original_response == reshaped_6_4
      ))
    }
    
    # Q6_5_A
    reshaped_6_5_A <- df_subject_reshaped |> 
      filter(iteration == iter) |> 
      select(Q6_5_A) |> 
      pull()
    original_response <- df_subject_original[[paste0("I_", iter, "_Q6_5_A")]]
    if (is.na(original_response) && is.na(reshaped_6_5_A)) {
      print(paste("Iteration", iter, "Q6_5_A:", TRUE))
    } else {
      print(paste(
        "Iteration",
        iter, 
        "Q6_5_A:", 
        original_response == reshaped_6_5_A
      ))
    }
  }
}
```

All looks good with the response reshaping.

Now we can join the two data frames back together:

```{r}
joined <- left_join(
  reshaped_vars,
  reshaped_responses,
  by = c("SbjNum", "iteration")
)
```

I will now convert the randomization variables and Q6_4 to factor, and Q6_5_A to
numeric.

```{r}
joined <- joined |> 
  mutate(
    across(contains("Q_7"), ~ factor(.x)),
    Q6_4 =  factor(Q6_4,
                   levels = c(
                     "1. Fully sufficient",
                     "2. Somewhat sufficient",
                     "3. Somewhat insufficient but better than nothing",
                     "4. Totally insufficient"
                   ),
                   labels = c(
                     "1. Fully sufficient",
                     "2. Somewhat sufficient",
                     "3. Somewhat insufficient",
                     "4. Totally insufficient"
                   ),
                   ordered = TRUE),
    Q6_5_A = as.numeric(Q6_5_A)
 )
```

I like to make a "final copy" of the data after I'm done.

```{r}
cleaned_data <- joined
```


## Analysis

I think I can now run a couple basic analyses.
First let's look at the distribution of response variables:

```{r}
table(cleaned_data$Q6_4, useNA = "always")
```

A plurality of respondents answered "4. Totally insufficient", with the other
options quite a bit lower.

```{r}
table(cleaned_data$Q6_5_A, useNA = "always")
```

Yes has more responses than No, but there are a large number of NAs.

## Ordered logit without respondent effects

I will now look at an ordered logit regression for Q6_4.
This model does not control for subject.

Please note, due to the length of the response values the tables below are
extremely wide. You can scroll side to side to show the point estimates and
other relevant statistics.

```{r}
q6_4_logit <- MASS::polr(Q6_4 ~ Q_75 + Q_76 + Q_77 + Q_78, 
                         data = cleaned_data, 
                         method = "logistic",
                         Hess = TRUE)

# Compute Odds Ratios and Confidence Intervals
exp(cbind(OR = coef(q6_4_logit), confint(q6_4_logit)))
```

It looks like Q75, the messaging around the loss, doesn't contain significant
effects across feature levels at the 0.05 level.

Note that Q6_4 is ordered such that lower values indicate greater sufficiency.
I can reverse that if we want.
So in looking at the data, an odds ratio below 1 indicates a higher degree of
sufficiency.

There are a number of feature levels where the confidence interval does not
contain 1:
- Q76: One-time payment of 3 million Dinars: OR 0.771
- Q76: One-time payment of 30 million Dinars: OR 0.321
- Q77: No evidence of wrongdoing: 0.830
- Q78: Public memorial: 0.713
- Q78: New schools: 0.643

I can cluster standard errors by respondent using the *sandwich* package:

```{r}
coeftest(q6_4_logit, vcov = vcovCL(q6_4_logit, cluster = ~ SbjNum))
```

With the clustered standard errors, we still see significant results at the 0.05 
level for:
- Q76: 3 million Dinars
- Q76: 30 million Dinars
- Q78: Public memorial
- Q78: New schools
Note that Q77: no evidence of wrongdoing is close to p < 0.05.

## Binary logit for Q6_5_A

Would the respondent accept the offer of compensation?

First, exclude randomizations that do not include compensation, and re-label the
factor

```{r}
compensation_rows <- cleaned_data |> 
  filter(Q_76 != "Unfortunately, we cannot offer you any monetary compensation for your loss.") |> 
  mutate(Q_76 = factor(
    Q_76,
    levels = c(
      "We can offer you a one-time payment of 3 million Iraqi Dinars as compensation for your loss.",
      "We can offer you a one-time payment of 30 million Iraqi Dinars as compensation for your loss."
      )
    )
    )

q6_5_a_logit <- glm(Q6_5_A ~ Q_75 + Q_76 + Q_77 + Q_78, 
                    data = compensation_rows,
                    family = binomial(link = "logit"))
summary(q6_5_a_logit)
```

Clustered SEs:

```{r}
coeftest(q6_5_a_logit, vcov = vcovCL(q6_5_a_logit, cluster = ~ SbjNum))

```

For this binary response, the only significant effect is the 30 million Dinars.
